---
title: "ecurve Primer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecurve Primer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction and Background

This document provides an introduction to `ecurve`, an R library that implements the Enhanced Standard Curve (ESC) model for constructing qPCR "standard curves".

The remainder of this section provides a brief description of the ESC model, while later sections illustrate the application of the model, through the `ecurve` package, to an example qPCR dataset. The same dataset will also be analyzed using established standard curve methods, in order to show the contrast between the two approaches.

The ESC model is a hierarchical model that attempts to capture the randomness inherent in both the number of gene copies from a sample that end up in a particular well and in the $C_q$ value that results from running qPCR on a well with a given initial number of gene copies. 

The initial number of gene copies in a given well, $N_0$, is modeled as a Poisson distribution 
$$N_0 \sim Poisson(\lambda)$$ 
where $\lambda$ is the concentration of the gene in the sample from which the well was drawn. 

If $N_0 = 0$, this is assumed to lead to a non-detect. Otherwise, as in traditional standard curve modelling, it is assumed that the $C_q$ value obtained during qPCR has a linear relationship with $\log N_0$, but for the ESC model we additionally assume that the measured $C_q$ values vary around this theoretical value according to a normal distribution with variance $\sigma^2$ independent of $N_0$:

$$C_q \sim N(\alpha + \beta \log N_0, \sigma^2)$$ 
Together, these assumptions fully determine the expected distribution of $C_q$ values for any sample concentration, and hence fully specify the model. The `ecurve` package provides mechanisms for fitting the model based on qPCR data from samples of known concentration via maximum likelihood estimation (MLE) of the (a priori) unknown parameters $\alpha$, $\beta$ and $\sigma$.

The MLE is performed using the `esc_mle()` function, for estimating the concentration of an unknown sample from qPCR data and a fitted model via the `conc_mle` function, and for using Bayesian methods to generate a credible interval to quantify the uncertainty in such a concentration estimate via the `conc_interval()` function. 

The functions `plot_esc_data()`, `plot_esc_model()` and `plot_conc_int()` are also provided to help visualize the results of these analyses.

## Data

The data used for this example analysis is located in two csv files: `esc-fit-data.csv` and `esc_test-data.csv`. 

The dataset `esc-fit-data.csv` contains data for fitting the model, and contains two columns, a `concentrations` column containing known sample concentrations, and a `cqs` column containing the corresponding $C_q$ value obtained in a qPCR run. 

The dataset `esc-test-data.csv` contains data for illustrating concentration estimation from other samples (not used in the model fit). It contains two columns, a `sample` column containing a numerical index of the sample for which we wish to know the concentration (due to the presence of technical replicates, these are not unique), and  `cqs` column containing the corresponding $C_q$ value obtained in a qPCR run. 

We now proceed to load the data:
```{r}
fit.data  <- read.csv("esc-fit-data.csv")
test.data <- read.csv("esc-test-data.csv")
```
In the datasets, non-detects are encoded by a $C_q$ value of "ND", which also results in the cqs values being interpreted as strings. However, the `ecurve` package requires the cqs values to be numeric with non-detects coded as `NaN`, so the data must be modified to fit this standard before the analysis is run.
```{r}
fit.data$cqs[which(fit.data$cqs == "ND")] <- NaN
fit.data$cqs <- as.numeric(fit.data$cqs)
test.data$cqs[which(test.data$cqs == "ND")] <- NaN
test.data$cqs <- as.numeric(test.data$cqs)
```

## Model Fitting

To fit the ESC model to the data, we use the `esc_mle()` function. This function takes a dataframe containing data on the known concentrations and $C_q$ values used to calibrate the model, as well as an optional parameter `CI` that controls the width of the credible interval displayed when plotting the model (defaults to 95%). 

The model is fit using MLE, and the function returns an object containing the fitted model parameters, `eff`, an estimate of the qPCR efficiency, `data`, a copy of the original data, and additional information to facilitate plotting.

First, let's make sure the `ecurve` library is loaded:
```{r load_library }
library(ecurve)
```

Now, let's fit the ESC model to the data

```{r model_fit, warning=FALSE}
model <- esc_mle(fit.data, CI = 0.95)
```

The maximum likelihood estimations for the ESC model parameter are:
```{r MLE_results}
model$intercept
model$slope
model$sigma
model$eff
```

As such, these MLE estimates are not directly useful. We will see how this fitted model will be used to directly estimate concentrations form $C_q$ values measured from new samples.

The model can be visualized by calling `plot_esc_model()`, which takes a fitted esc model object and produces a plot of $C_q$ value versus the log concentration. The black points represent the datapoints used to fit the model, and the blue line shows the median $C_q$ value for each concentration as estimated by the model. Finally, the light blue area indicates the 95\% probability interval (or any other percentage, as specified by the input parameter `CI` when fitting the model) for the $C_q$ values given the fitted model prarmeters. That is, if the fitted model is indeed an accurate reflection of the data generating mechanism, we would expect about 95\% of the data points to fall within this region. The plot is returned by the function as a `ggplot` object.
```{r plot_esc_model, fig.width=6}
plot_esc_model(model)
```

Note that, with the exception of a few outliers, all of the datapoints lie within the light blue region, indicating that the model is accurately capturing the variability in the data.

For contrast, we also illustrate fitting a simple linear model regressing $C_q$ on log concentration as in traditional standard curve analysis. To visualize this model, we use the `plot_esc_data()` and then manually add the regression line (using `ggplot` built-in regression):
```{r linear_regression, warning=FALSE, fig.width=6}
plot_esc_data(fit.data) + 
  ggplot2::geom_smooth(method = 'lm', formula = 'y~x', se = FALSE)
```

Note the difference of the functional relationship at low concentrations. The naive linear regression cannot account for the greater spread in $C_q$ values observed at lower concentrations, or indeed meaningfully quantify that spread at all.



## Concentration Estimation

To illustrate concentration estimation using the `ecurve` package, we will focus on samples 3, 7, and 8 from the test data. First, we must isolate the $C_q$ values associated with the technical replicates from each sample:
```{r extract_samples_example, warning=FALSE}
samp3_cqs <- test.data$cqs[which(test.data$sample == 3)]
samp7_cqs <- test.data$cqs[which(test.data$sample == 7)]
samp8_cqs <- test.data$cqs[which(test.data$sample == 8)]
samp3_cqs
samp7_cqs
samp8_cqs
```

Notice that sample 7 contains multiple non-detects, and sample 8 is in fact all non-detects. However, no special measures need to be taken to account for this, as the ESC model smoothly incorporates non-detects into the concentration estimation procedure. 
Concentration estimates can be obtained using the `conc_mle()` function, produces them using MLE given a set of $C_q$ values from technical replicates and the model fitted in the previous section:

```{r, warning=FALSE}
conc_mle(samp3_cqs, model = model)
conc_mle(samp7_cqs, model = model)
conc_mle(samp8_cqs, model = model)
```

The function `conc_mle()` provides the "best" estimates, according to the MLE method (ie the parameter value that maximisethe likelihood of the data). 
We can also quantify the uncertainty in these estmates using the `conc_interval()` function. This function uses numerical Bayesian estimation to compute the posterior concentration distribution given the $C_q$ data and fitted model, assuming a uniform prior, and construct a credible interval for the concentration. 
The credible level of the interval can be controlled using the optional `level` parameter. The function returns an object containing a list named `interval` containing the limits of the interval as well as the MLE concentration. It also contains a data frame giving information on the posterior distribution for use in visualizing the interval
```{r concentration_intervals}
samp3_int <- conc_interval(samp3_cqs, model = model)
samp7_int <- conc_interval(samp7_cqs, model = model)
samp8_int <- conc_interval(samp8_cqs, model = model)
#Note: credible level for intervals is 95% by default
samp3_int$interval
samp7_int$interval
samp8_int$interval
```
These intervals can also be visualized using the `plot_conc_int()` function, which takes a constructed credible interval and plots either the probability density function (input parameter `type = "pdf"`) or the cumulative density function (`type = "cdf"`) of the posterior distribution of the concentration, as specified by the parameter type, with the MLE and interval endpoints marked. As with other visualization functions, the plot is returned as a `ggplot` object. For brievity, this is only illustrated for sample 7.
```{r plot_concetration_pdf, fig.width=8}
plot_conc_int(samp7_int, type = "pdf")
plot_conc_int(samp7_int, type = "cdf")
```


For comparison, we also compute concentration estimates using the "traditional" method based on the linear model fitted earlier. We plot the "traditional" estimates with the ESC estimates: 

```{r, echo=FALSE}
# Naive linear model:

fit.data$logconc = log(fit.data$concentrations)

lin_mod <- lm(cqs ~ logconc, data = fit.data)
intercept.m = lin_mod$coef[1]
slope.m     = lin_mod$coef[2]

# Calculate confidence intervals of the linear regression
lin_ci = confint(lin_mod)
intercept.lo = lin_ci[1,1]
intercept.hi = lin_ci[1,2]
slope.hi     = lin_ci[2,1]
slope.lo     = lin_ci[2,2]

newdata = data.frame(logconc = seq(-2.4, 5.3, length.out=100))

p = predict.lm(lin_mod, 
               newdata = newdata,
               se.fit=TRUE, interval="confidence", level=0.95)

foo = cbind(p$fit, newdata)
foo$conc = exp(foo$logconc)

# Inferred concentrations:
samp3_mean_cq <- mean(samp3_cqs, na.rm = TRUE)
samp7_mean_cq <- mean(samp7_cqs, na.rm = TRUE)
samp3_conc    <- unname(exp((samp3_mean_cq - intercept.m)/slope.m))
samp7_conc    <- unname(exp((samp7_mean_cq - intercept.m)/slope.m))

get_lin_ci <- function(mcq, foo) {
  idx     = which.min(abs(foo$fit - mcq))
  idx.lwr = which.min(abs(foo$lwr - mcq))
  idx.upr = which.min(abs(foo$upr - mcq))
  res = c(
    lwr = foo$conc[idx.lwr],
    m  = foo$conc[idx],
    upr = foo$conc[idx.upr])
  return(res)
}

lin.ci.3 = get_lin_ci(mcq = samp3_mean_cq, foo)
lin.ci.7 = get_lin_ci(mcq = samp7_mean_cq, foo)

```


``` {r, plot_comparison, echo = FALSE, fig.width = 7}

col.segm = 'red3'
alpha.segm = 0.5
lw.segm = 3

g = plot_esc_model(model) + 
  ggplot2::geom_hline(yintercept = c(samp3_mean_cq, samp7_mean_cq),
                      linetype = 'dashed', color = 'grey50')+
  # -- ESC model
  ggplot2::annotate("segment", 
                    x = samp3_int$interval$lower, xend = samp3_int$interval$upper,
                    y = samp3_mean_cq, yend = samp3_mean_cq, 
                    color="black", linewidth = lw.segm/2, alpha = alpha.segm)+
  ggplot2::annotate("segment", 
                    x = samp7_int$interval$lower, xend = samp7_int$interval$upper,
                    y = samp7_mean_cq, yend = samp7_mean_cq, 
                    color="black", linewidth = lw.segm/2, alpha = alpha.segm)+
  # -- linear model
  ggplot2::annotate("segment", 
                    x = lin.ci.3['lwr'], xend = lin.ci.3['upr'],
                    y = samp3_mean_cq, yend = samp3_mean_cq, 
                    color=col.segm, linewidth = lw.segm, alpha = alpha.segm)+
  ggplot2::annotate("segment", 
                    x = lin.ci.7['lwr'], xend = lin.ci.7['upr'],
                    y = samp7_mean_cq, yend = samp7_mean_cq, 
                    color=col.segm, linewidth = lw.segm, alpha = alpha.segm)+
  ggplot2::labs(title = 'Comparison ESC vs. traditional linear regression', subtitle = NULL)
plot(g)

```

The horizontal dashed line shows the mean $C_q$ values, for samples 3 and 7, used for the traditional linear regression and the red segments represents the 95% confidence intervals for the associated concentrations with that model.

The dark gray segments represent the 95% credible intervals for samples 3 and 7 using the ESC model. 

While the central estimates are similar, we note that their respective uncertainty (red and gray segments) are markedly different: a much larger uncertainty is implied by the ESC model. 
For high $C_q$ values, that may be more representative of reality (i.e. the traditional linear regression may be over-confident at low concentrations).

We also note that no estimate can be generated by the traditional linear regression for sample 8 at all since all the replicates are non-detects whereas it is possible for the ESC model.
